Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-10-14 paper/main.tex
polishing of 3.1

The formula $\Phi_{1,C,K}$ built with $\mcycle_1(\vec{e})$ actually
says that for every nonempty set $T\subseteq\delta_S$ whenere $S$ is
an SCC, $T$ satisfies $\varphi$ if and only if $\rem(T)$ satisfies
$\psi$. Note that the formula is not affected by the placement of
transitions in the automaton. \todo{Tady to predelat, aby to
  reflektovalo $T\subseteq\delta_S$.}  The only aspect of a transition
$t$ reflected by the formula is its set of marks $\mks(t)$. Hence, we
do not have to distinguish between transitions with the same sets of
marks.

Formally, we define an equivalence ${\sim}\subseteq\delta\times\delta$
on transitions such that $t_1\sim t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma}
  Assume that there is a function $\rem$ and a formula $\psi$ such that 
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects % the equivalence
  $\sim$ (i.e.,
  if $t_1\sim t_2$ then $\rem'(t_1)=\rem'(t_2)$) and
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of $\sim$. For
  every transition $t\in\delta$, by $\overline{t}$ we denote the
  selected transition equivalent to $t$. We define the function
  $\rem'$ as $\rem'(t)=\rem(\overline{t})$. Clearly, $\rem'$ respects
  the equivalence $\sim$. It remains to show that \eqref{eq:lem2}
  holds for $\rem'$ and $\psi$.

  Let $T\subseteq\delta$ be an arbitrary nonempty set. We construct
  the set $\overline{T}=\{\overline{t}\mid t\in T\}$. As
  $\mks(t)=\mks(\overline{t})$ for all transitions, we get
  $\mks(T)=\mks(\overline{T})$ and thus
  \[
    T~\msat~\varphi~\iff~\overline{T}~\msat~\varphi.
  \]
  Now we apply $\eqref{eq:lem1}$ on $\overline{T}$ to get that
  \[
    \overline{T}~\msat~\varphi~\iff~\rem(\overline{T})~\msat~\psi.
  \]
  Finally, the definition of $\rem'$ implies that
  $\rem'(T)=\rem(\overline{T})$ and thus
  \[
    \rem(\overline{T})~\msat~\psi~\iff~\rem'(T)~\msat~\psi.
  \]
  Altogether, we obtain
  \[
    \setlength{\arraycolsep}{0pt}
    \begin{array}{rl}
      T~\msat~\varphi~ 
      & \iff~\overline{T}~\msat~\varphi \\
      & \iff~\rem(\overline{T})~\msat~\psi\\
      & \iff~\rem'(T)~\msat~\psi
    \end{array}
  \]
  which proves that \eqref{eq:lem2} holds for $\rem'$ and $\psi$.\qed
\end{proof}

The lemma suggests the following simplification of the whole formula
$\Phi_{1,C,K}$ built with $\mcycle_1(\vec{e})$. Before we build the
formula, we compute the equivalence $\sim$ on automata transitions and
temporarily remove all transitions except one of each equivalence class. Then we
build the formula $\Phi_{1,C,K}$ for the pruned automaton. The more
transitions we removed, the shorter formula with less $e_t$ variables
we obtain. If the formula $\Phi_{1,C,K}$ for the pruned automaton is
satisfiable, we derive $\rem$ and $\psi$ from its model and extend % use them
% with the original automaton. We only need to extend
$\rem$ to all transitions of the original automaton such that
it changes the acceptance marks on all equivalent transitions in the
same way. In the following, we use this simplification whenever
$\Phi_{1,C,K}$ is employed.

@@

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transtitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not holds for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradition that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.\qed
\end{proof}

@@

The second optimization is specific for $\Phi_{3,C,K}$. In
construction of $\mcycle_3(\vec{e})$, we replace the subformula
$\bigvee_{t\in\delta}e_t$ enforcing the nonemptiness of $\Te$ by
$\mcycle_2(\vec{e})$. This modification does not change the overall
semantics of $\mcycle_3(\vec{e})$, but our prelimnary experiments
showed that QBF solvers can often solve the modified formula
$\Phi_{3,C,K}$ faster.

>>>>

The formula $\Phi_{1,C,K}$ built with $\mcycle_1(\vec{e})$ says that
for every nonempty set $T\subseteq\delta_S$ whenere $S$ is an SCC, $T$
satisfies $\varphi$ if and only if $\rem(T)$ satisfies $\psi$. Note
that the only aspects of a transition $t$ reflected by the formula are
its set of marks $\mks(t)$ and its affiliation to an SCC.
% In partcular, it is not affected by its placement inside the
% corresponding SCCs. The only aspect of a transition $t$ reflected by
% the formula is its set of marks $\mks(t)$.
Hence, we do not have to distinguish between transitions that are
affiliated to the same SCC an have the same sets of marks.

Let us now fix an SCC $S$. We define an equivalence
${\sim_S}\subseteq\delta_S\times\delta_S$ on transitions % in $\delta_S$
such that $t_1\sim_S t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma} Assume that there is a function $\rem$ and a formula
  $\psi$ such that
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects % the equivalence
  $\sim_S$ (i.e.,
  if $t_1\sim_S t_2$ then $\rem'(t_1)=\rem'(t_2)$) and
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of~$\sim_S$. For
  every transition $t\in\delta_S$, by $\overline{t}$ we denote the
  selected transition equivalent to $t$ and we define the function
  $\rem'$ as $\rem'(t)=\rem(\overline{t})$. Note that we do not need
  to discuss the value of $\rem'$ on transitions outside $\delta_S$ as
  it is not relevant for the lemma. Clearly, $\rem'$ respects the
  equivalence $\sim_S$. It remains to show that \eqref{eq:lem2} holds
  for $\rem'$ and $\psi$.

  Let $T\subseteq\delta_S$ be a % an arbitrary
  nonempty set. We construct
  the set $\overline{T}=\{\overline{t}\mid t\in T\}$. As
  $\mks(t)=\mks(\overline{t})$ for all transitions of $\delta_S$, we
  get $\mks(T)=\mks(\overline{T})$ and thus
  \[
    T~\msat~\varphi~\iff~\overline{T}~\msat~\varphi.
  \]
  Now we apply $\eqref{eq:lem1}$ on $\overline{T}$ to get that
  \[
    \overline{T}~\msat~\varphi~\iff~\rem(\overline{T})~\msat~\psi.
  \]
  Finally, the definition of $\rem'$ implies that
  $\rem'(T)=\rem(\overline{T})$ and thus
  \[
    \rem(\overline{T})~\msat~\psi~\iff~\rem'(T)~\msat~\psi.
  \]
  Altogether, we obtain
  \[
    \setlength{\arraycolsep}{0pt}
    \begin{array}{rl}
      T~\msat~\varphi~ 
      & \iff~\overline{T}~\msat~\varphi \\
      & \iff~\rem(\overline{T})~\msat~\psi\\
      & \iff~\rem'(T)~\msat~\psi
    \end{array}
  \]
  which proves that \eqref{eq:lem2} holds for $\rem'$ and $\psi$.\qed
\end{proof}

The lemma suggests the following simplification of the whole formula
$\Phi_{1,C,K}$ built with $\mcycle_1(\vec{e})$. Before we build the
formula, we compute the equivalences $\sim_S$ for all SCCs and
temporarily remove all transitions affilited to SCCs except one of
each equivalence class. Then we build the formula $\Phi_{1,C,K}$ for
the pruned automaton. The more transitions we removed, the shorter
formula with less $e_t$ variables we obtain. If the formula
$\Phi_{1,C,K}$ for the pruned automaton is satisfiable, we derive
$\rem$ and $\psi$ from its model and extend % use them
% with the original automaton. We only need to extend
$\rem$ to all transitions of the original automaton such that
it changes the acceptance marks on all equivalent transitions in the
same way. In the following, we use this simplification whenever
$\Phi_{1,C,K}$ is employed.

@@

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transtitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not holds for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradition that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
/  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.\qed
\end{proof}

@@

The second optimization is specific for $\Phi_{3,C,K}$. In
the construction of $\mcycle_3(\vec{e})$, we replace the subformula
$\bigvee_{t\in\delta}e_t$ enforcing the nonemptiness of $\Te$ by
$\mcycle_2(\vec{e})$. This modification does not change the overall
semantics of $\mcycle_3(\vec{e})$, but our prelimnary experiments
showed that QBF solvers can often solve the modified formula
$\Phi_{3,C,K}$ faster.
